### 2024-01-17 parts/starfree.tex
ultimate polynomiality is a lie

The goal of the upcoming section is to provide a notion of residual transducer
for \kl{commutative} \kl{$\Nat$-polyregular functions}.

\begin{fact}
    For every $K \in \Nat$, and every $n \in \Nat$,
    $\Nat^n$ is a \kl{well-quasi-ordering}
    when endowed with the quasi-ordering $\vec{u} \leq_K \vec{v}$
    if and only if $\vec{u} = \vec{v}$ or
    $\vec{u}$ is smaller than $\vec{v}$ for the product ordering and
    $\sum_{i = 1}^n  (\vec{v} - \vec{u})_i \geq K$.
\end{fact}
\begin{proof}
    First, $\Nat^n$ with the product ordering is a \kl{well-quasi-ordering}.
    Therefore, given a sequence $\seqof{\vec{u}_i}{i \in \Nat}$,
    one can extract an infinite monotone subsequence for
    the product ordering.
    Either there exists $i < j$ such that $\vec{u}_i = \vec{u}_j$
    in the subsequence, in which case $\vec{u}_i \leq \vec{u}_j$
    and the original sequence was good. Or for all $i < j$,
    we have $\vec{u}_i$ that is strictly smaller than 
    $\vec{u}_j$ for the product ordering. As a consequence,
    $\vec{u}_0 \leq \vec{u}_K$.
\end{proof}

@@

\begin{fact}
    Let $P \in \CorrectPoly[X,\vec{Y}]$, and $K,L \in \Nat$,
    then $P(X+L, \vec{Y}) - P(X+K,\vec{Y}) \in \CorrectPoly$
    whenever $L \geq K$.
\end{fact}
\begin{proof}
    We can write
    $P = \sum_{i = 0}^n P_i(\vec{Y}) X^i$, with $P_i \in \Nat[\vec{Y}]$.
    Now,
    $P(X+L,\vec{Y}) - P(X+K,\vec{Y})$
    is precisely
    $\sum_{i = 1}^n P_i(\vec{Y}) (L-K) \left[ \sum_{j + l + 1= i} (X+L)^j (X+K)^l \right]$.
    Now, $P_i(\vec{Y})$
    remaining with positive coefficients.
\end{proof}

@@

@@

\begin{defintion}
    Residual $\Nat$-transducer.
\end{definition}

\begin{lemma}
    Computable 
\end{lemma}

\begin{theorem}
    If and only if this one is counter-free.
\end{theorem}

>>>>

The goal of the upcoming section is to provide a notion of residual transducer
for \kl{commutative} \kl{$\Nat$-polyregular functions}.
This will be based on another characterization of $\CorrectPoly$
using partial derivatives. Namely, we define

\begin{definition}
    For every polynomial $P \in \Rel[X \vec{Y}]$, we define 
    the partial derivative with respect to $X$ as follows:
    \begin{equation*}
        \partial_X P \defined P(X + 1, \vec{Y}) - P(X, \vec{Y}) \quad .
    \end{equation*}
\end{definition}

\begin{fact}
    Let $P \in \CorrectPoly[X_1, \dots, X_n]$, then
    for all $1 \leq i \leq n$,
    $\partial_{X_i} P \in \CorrectPoly$.
\end{fact}
\begin{proof}
    We prove the result by induction on the number of variables.
    Let $\nu \colon \vec{X} \topartial \Nat$.
    Two cases, either $X_i \in \dom(\nu)$,
    in which case, we remark that
    we can split $\nu$ into two valuations,
    and that coefficients are good.

    Otherwise, evaluation commutes with partial,
    and we conclude by induction hypothesis.
\end{proof}

\begin{remark}
    The converse does not hold,
    for instance $P(X) \defined X - 1$ has all its derivatives
    that are in $\CorrectPoly$, but is not itself in $\CorrectPoly$.
\end{remark}

@@

\begin{fact}
    Let $P \in \CorrectPoly[X,\vec{Y}]$, and $K,L \in \Nat$,
    then $P(X+L, \vec{Y}) - P(X+K,\vec{Y}) \in \CorrectPoly$
    whenever $L \geq K$.
\end{fact}
\begin{proof}
    We can write
    $P = \sum_{i = 0}^n P_i(\vec{Y}) X^i$, with $P_i \in \Nat[\vec{Y}]$.
    Now,
    $P(X+L,\vec{Y}) - P(X+K,\vec{Y})$
    is precisely
    $\sum_{i = 1}^n P_i(\vec{Y}) (L-K) \left[ \sum_{j + l + 1= i} (X+L)^j (X+K)^l \right]$.
    Now, $P_i(\vec{Y})$
    remaining with positive coefficients.
\end{proof}
\begin{fact}
    For every $K \in \Nat$, and every $n \in \Nat$,
    $\Nat^n$ is a \kl{well-quasi-ordering}
    when endowed with the quasi-ordering $\vec{u} \leq_K \vec{v}$
    if and only if $\vec{u} = \vec{v}$ or
    $\vec{u}$ is smaller than $\vec{v}$ for the product ordering and
    $\sum_{i = 1}^n  (\vec{v} - \vec{u})_i \geq K$.
\end{fact}
\begin{proof}
    First, $\Nat^n$ with the product ordering is a \kl{well-quasi-ordering}.
    Therefore, given a sequence $\seqof{\vec{u}_i}{i \in \Nat}$,
    one can extract an infinite monotone subsequence for
    the product ordering.
    Either there exists $i < j$ such that $\vec{u}_i = \vec{u}_j$
    in the subsequence, in which case $\vec{u}_i \leq \vec{u}_j$
    and the original sequence was good. Or for all $i < j$,
    we have $\vec{u}_i$ that is strictly smaller than 
    $\vec{u}_j$ for the product ordering. As a consequence,
    $\vec{u}_0 \leq \vec{u}_K$.
\end{proof}

@@

\begin{definition}
    Residual $\Nat$-transducer.
\end{definition}

\begin{lemma}
    If $P \in \CorrectPoly$, then it is computable
    by a $\Nat$-residual transducer.
\end{lemma}

\begin{theorem}
    The following are equivalent for \kl{commutative}.
    \begin{itemize}
        \item $f$ is in $\NSF$,
        \item The residual automata of $f$ has no counters
    \end{itemize}
\end{theorem}

\section{Beyond commutativity}

\begin{lemma}
    Let $f \in \NPoly[k]$,
    and $h \colon \Nat^p \to \Sigma^*$ be a polyregular function.
    Then,
    $f \circ h$ is a \kl{commutative} \kl{polyregular function},
    and
    the set $\setof{ f(u \cdot) \circ h }{u \in \Sigma^* }$
    is a \kl{well-quasi-ordering}
    for the order
    $g_1 \leq g_2$ iff 
    $g_2 - g_1 \in \NPoly[k-1]$.
\end{lemma}

\begin{claim}
    Let $k \in \Nat$,
    there exists a finite family $\mathcal{F}_k$
    of \kl{commutative} \kl{polyregular functions}
    such that the following are equivalent
    for all $f \in \ZPoly[k]$:
    \begin{itemize}
        \item $f \in \NPoly[k]$,
        \item For all $h \in \mathcal{F}_k$,
            $(f \circ h)$ is a \kl{commutative}
            \kl{$\Nat$-polyregular function}.
    \end{itemize}
\end{claim}

@@

\begin{lemma}
    If $f \in \NPoly[k]$, then the algorithm computes
    a residual transducer of $f$.
\end{lemma}

\begin{theorem}
    The following are equivalent for $f \in \NPoly$.
    \begin{itemize}
        \item The function $f \in \NSF$,
        \item The residual transduucer of $f$ has no counters,
        \item For all commutative polyregular functions $h$,
            $f \circ h$ is a \kl{commutative}
            \kl{star-free $\Nat$-polyregular function},
        \item $f$ is \kl{ultimately polynomial}.
    \end{itemize}
\end{theorem}
